---
title: "Ggplot Asssignment"
author: "Charlene Fasuyi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r Libraries used}
library(scales)       # Used to customize and format scales, including log scales and axis labels.
library(rnaturalearth) # Provides map data for countries, used for creating the base world map.
library(rnaturalearthdata) # Contains additional geographical data for mapping.
library(gridExtra)    # Used to arrange multiple ggplot objects for comparison.
library(forcats)      # Used for reordering and manipulating factor levels in categorical data.
library(plotly)       # Used to create interactive plots and convert ggplot visualizations to interactive Plotly plots.
library(tidyverse)    # Used for data wrangling (e.g., dplyr, tidyr) and visualization (e.g., ggplot2).
```


```{r Importing data table}
emission_data <- read_csv("Files/MASTER.csv")
```


## World Map data. 
```{r Importing World countries data}
# Load world map data
world <- ne_countries(scale = "medium", returnclass = "sf")
```


```{r Combined emissions and world map data table}
# Joined emissions and world map data by country code. 
map_data <- world %>%
              left_join(emission_data, 
                        by = c("brk_a3" = "ISO.alpha-3"))
```


```{r (2021) Wolrd emission data preparation}
# Filtered data for years 1990 to 2020.
map_data <- map_data %>% filter(Year == 2021)
```


```{r Map of Total CO2 emissions by country}
world_emissions_map <-  ggplot(data = map_data
                        ) +
                        geom_sf(
                               aes(
                                  fill = Total.CO2, # Colours map according to CO2 emissions
                                  text = paste("Country: ", Country,  # Formats hover text to include both country name and total CO2 emissions
                                                "\nTotal CO2 Emissions: ", format(
                                                                                 Total.CO2, # Formats CO2 emissions with commas as thousands separators and removes scientific notation.
                                                                                 big.mark = ",", 
                                                                                 scientific = FALSE)
                                  )
                                  ), 
                               color = "black", size = 0.1
                        ) + # Makes country boundaries black
                        scale_fill_gradientn(      # Customize colour to personal preference
                                            colors = c("#FFF5E1", 
                                                       "#FFC87C", 
                                                       "#FF8C42", 
                                                       "#FF5722", 
                                                       "#D81B1B"),# Smooth gradient using sequential scheme to signifiy increase
                                            breaks = seq(min(map_data$Total.CO2, 
                                                             na.rm = TRUE), # Ensures legend starts with smallest total emissions.
                                                         max(map_data$Total.CO2, 
                                                             na.rm = TRUE), # Ensures legend ends with largest total emissions.
                                                         length.out = 5),  # Ensure 5 levels on legend
                                            na.value = "grey80"
                        ) +
                        theme_minimal() 
```


```{r Ggplot map to plotly map}
#Convert the ggplot to an interactive Plotly plot
plotly_map <- ggplotly(world_emissions_map, tooltip = "text")

```


```{r Interactive plotly emissions world map}
# Adding the title and subtitle to map graph
plotly_map <- plotly_map %>%
                     layout(
                           title = list(
                                       text = "Global CO<sub>2</sub> Emissions (2021)<br><sup>Measured in millions of tons</sup>",
                                       y = 0.86,
                                       xanchor = "center"
                                       )
                           )


```


```{r Display World Emissions Graph}
plotly_map
```



## Industry Emissions
```{r Industry emissions data preparation}
# Group the data by 'Year' and summarize emissions by industry
industry_emissions <- emission_data %>% 
                               group_by(Year) %>%  # Group the data by 'Year' to summarize emissions for each year
                               summarise(       # Calculate total emissions for each industry (ignoring NAs)
                                        Coal.CO2 = sum(Coal.CO2, na.rm = TRUE),  
                                        Oil.CO2 = sum(Oil.CO2, na.rm = TRUE),    
                                        Gas.CO2 = sum(Gas.CO2, na.rm = TRUE),    
                                        Cement.CO2 = sum(Cement.CO2, na.rm = TRUE),  
                                        Flaring.CO2 = sum(Flaring.CO2, na.rm = TRUE),  
                                        Other.CO2 = sum(Other.CO2, na.rm = TRUE)  
                               ) %>% 
                               ungroup(   # Return data to data frame form.
                               ) %>%  
                               pivot_longer(
                                           cols = contains("CO2"),  # Select columns that contain "CO2" in their names for pivoting
                                           names_to = "Industry",  # Create a new column 'Industry' for the industry names
                                           values_to = "Emissions"  # Create a new column 'Emissions' for the emission values
                               ) %>% 
                               mutate(Industry = str_replace_all(Industry, ".CO2", "") # Clean up the 'Industry' column by removing the '.CO2' suffix
                               )
```


```{r Reordering Factor Levels}
#Reordered 'Industry' factor levels for plot legend
industry_emissions$Industry <- fct_relevel(industry_emissions$Industry, 
                                          "Coal", "Oil", "Gas", "Cement", "Flaring", "Other")
```


```{r Color for industry graph}
#Define custom colors for each industry
industry_colors <- c(
                    "Coal" = "#D81B1B",  # Dark red for coal (most emissions)
                    "Oil" = "#FF5722",   # Orange for oil
                    "Gas" = "#FF8C42",   # Lighter orange for gas
                    "Cement" = "#FFB300",  # Yellow for cement
                    "Flaring" = "#FFC87C", # Light yellow-orange for flaring
                    "Other" = "#D3D3D3"    # Light gray for other industries
                    )
```


```{r Emissions per Industry line plot}
industry_line <- ggplot(
                       industry_emissions, aes(
                                              x = Year, 
                                              y = Emissions, 
                                              color = Industry)
                       ) +
                       geom_line(size = 1      # Plots line graph 
                       ) +  
                       scale_color_manual(values = industry_colors   # Use custom colors specified above
                       ) +  
                       labs(
                           color = "Industry", 
                           caption = "Data Source: Miller, A.(2024). Global Energy and Emissions Dataset (1980–2021)."
                       ) +
                       theme_minimal()   
  
```


```{r Plotly Industry graph creation}
#Convert the ggplot to an interactive Plotly plot
 plotly_industry <- ggplotly(industry_line)
```


```{r Industry plotly graph customisation}
# Adding the title, subtitle and axis labels 
 plotly_industry <-  plotly_industry %>%
                            layout(
                                  title = list( 
                                              text = "Global CO<sub>2</sub> Emissions per Industry(1980 - 2021)<br><sup>Measured in million tonnes</sup>", 
                                              x = 0.5,
                                              y = 0.95,
                                              xanchor = "center"
                                  ),
                                  yaxis = list(
                                              title = "CO<sub>2</sub> Emissions" # Adding subscript for CO2
                                  ),
                                  xaxis = list(
                                              title = "Year") 
                                  )

```


```{r Display Industry graph}
 plotly_industry
```


## Highest and Lowest Emitters per year

```{r Data preparation for high CO2 emitters}
emitters_high <-  emission_data %>%
                     select(Year, 
                            Country, 
                            Total_CO2 = Total.CO2)
```


```{r Identify the highest CO2 emissions per year}
emitters_high1 <- emitters_high %>% 
                           mutate(Total_CO2 = ifelse(Total_CO2 == 0, 
                                                      NA, 
                                                      Total_CO2) # Replace zero values in Total_CO2 column with NA
                                                    ) %>%  
                           group_by(Year) %>% 
                           summarise(Emission = max(Total_CO2, 
                                                     na.rm = TRUE) # Find maximum emissions for each year
                                                    ) %>%  
                           ungroup()
```


```{r Join high emitters data with original data to get country information}
emitters_high1 <-  emitters_high1 %>% 
                            left_join(emitters_high, 
                                      by = c("Emission" = "Total_CO2", "Year" = "Year"))
```


```{r Data preparation for low CO2 emitters}
emitters_low <- emission_data %>%
                         select(Year, 
                                Country, 
                                Total_CO2 = Total.CO2)
```


```{r Identify the lowest CO2 emissions per year}
emitters_low1 <- emitters_low %>% 
                          mutate(Total_CO2 = ifelse(Total_CO2 == 0, 
                                                    NA, 
                                                    Total_CO2) # Replace zero values in Total_CO2 column with NA
                                                   ) %>%  
                          group_by(Year) %>% 
                          summarise(Emission = min(Total_CO2, 
                                                   na.rm = TRUE) # Find minimum emissions for each year
                                                  ) %>%  
                          ungroup()
```


```{r Join low emitters data with original data to get country information}
emitters_low1 <- emitters_low1 %>% 
                          left_join(emitters_low, 
                                    by = c("Emission" = "Total_CO2", "Year" = "Year")
                                    )
```


```{r High emitters plot}
emitters_high1$Country <- as.factor(emitters_high1$Country)  # Convert 'Country' to factor for better plotting

high_emitters_plot <- ggplot(
                            emitters_high1, aes(
                                               x = Year, 
                                               y = Emission, 
                                               fill = Country)
                      ) +
                      geom_bar(
                              stat = "identity",  # Creates regular bar chart for high emitters
                              alpha = 0.7
                      ) + 
                      scale_fill_manual(
                                       values = c("#d94701", # Set custom colors for the plot
                                                  "#fd8d3c")
                      ) +  
                      labs(
                          title = expression("Highest CO"[2]~ "Emitters"), 
                          x = "Year",
                          y = expression("CO"[2]~ "Emissions (tons)")
                      ) +
                      theme(
                           panel.grid.major.y = element_line(colour = "grey90"), #Coloured gridlines for the y-axis
                           panel.background = element_rect(fill = "white")  # Every order chart element becomes white
                      ) 
```


```{r Low emitters plot}
low_emitters_plot <- ggplot(
                           emitters_low1, aes(
                                             x = Year, 
                                             y = Emission, 
                                             fill = Country)
                     ) +
                     geom_bar(
                             stat = "identity", # Plot stacked bar chart for low emitters
                             alpha = 0.7, 
                             position = "stack"
                     ) + 
                     scale_fill_manual(
                                      values = c("#deebf7", # Set custom colors for the plot
                                                 "#9ecae1", 
                                                 "#3182bd", 
                                                 "#2171b5")
                     ) +  
                     labs(
                         title = expression("Lowest CO"[2]~ "Emitters"),
                         x = "Year",
                         y =  expression("CO"[2]~ "Emissions (tons)")
                     ) +
                     scale_y_continuous(limits = c(0, 0.07)  # Limit y-axis scale for better visualization
                     ) +  
                     theme( 
                          panel.grid.major.y = element_line(colour = "grey90"), #Coloured gridlines for the y-axis
                          panel.background = element_rect(fill = "white")  # Every order chart element becomes white
                      ) 

```


```{r Arrange low and high emitters plots}
# Arrange the two plots vertically
grid.arrange(
            high_emitters_plot, 
            low_emitters_plot, 
            nrow = 2  # Arrange the plots in 2 rows
            )
```


## Scatterplot of Climate Change Impact vs Scatterplot of Green Energy Efforts

```{r Emission data for 2021}
# Filter the data for the year 2021
emission_2021 <- emission_data %>% filter(Year == 2021)
```


```{r Temp_emission_plot}
# Scatterplot: Total CO2 vs Temperature Change 

temp_emission_plot <- ggplot(
                            emission_2021, aes(
                                              x = Total.CO2, 
                                              y = Temp_Change, 
                                              color = Temp_Change, 
                                              fill = Temp_Change)
                      ) +
                      geom_point(
                                shape = 21,  # Shape with borders and fill
                                size = 3,    # Point size
                                color = "#3E2723",  # Border color
                                stroke = 1   # Border width
                      ) +
                      scale_y_continuous(
                                        breaks = c(-0.5, 0, 0.5, 1, 1.5, 2)  # Custom y-axis breaks
                      ) +
                      scale_x_continuous(
                                        trans = "log10",                    # Apply log scale to spread out points
                                        breaks = c(0, 1, 10, 100, 1000, 10000), # Custom x-axis breaks
                                        labels = comma                      # Format x-axis labels with commas
                      ) + 
                      scale_fill_distiller(
                                          palette = "Reds",                 # Gradient for the fill color
                                          direction = 1                     # Gradient direction
                      ) +  
                      labs(
                          title = expression("Total CO"[2] ~ "vs Temperature Change"),
                          x = expression("Total CO"[2] ~ "Emissions (Million Tonnes)"),
                          y = "Temperature Change (°C)",
                          fill = "Temp Change"
                      ) +
                      theme(
                           panel.grid.major.x = element_line(colour = "grey80"), # Major grid lines
                           panel.grid.major.y = element_line(colour = "grey90"),
                           panel.background = element_rect(fill = "white"),  # White background
                           plot.title = element_text(size = 15, face = "bold")  # Title customization
                      ) 
```


```{r Green_energy_plot}
# Scatterplot: Total CO2 vs Renewable Energy Adoption 

green_energy_plot <- ggplot(
                           emission_2021, aes(
                                             x = Total.CO2, 
                                             y = Renewables.and.other.Energy, 
                                             color = Renewables.and.other.Energy, 
                                             fill = Renewables.and.other.Energy)
                    ) +
                    geom_point(
                              shape = 21,  # Shape with borders and fill
                              size = 2,    # Point size
                              color = "#355E3B",  # Border color (dark green)
                              stroke = 1   # Border width
                    ) +
                    scale_y_continuous(
                                      breaks = c(0.5, 5, 10, 15, 20)  # Custom y-axis breaks    
                    ) +   
                    scale_x_log10(
                                 breaks = c(0, 1, 10, 100, 1000, 10000), # Log scale for x-axis
                                 labels = comma    # Format x-axis labels with commas
                    ) +
                    scale_fill_distiller(
                                        palette = "Greens", direction = 1  # Gradient for the fill color
                    ) +
                    labs( 
                        title = expression("Total CO"[2] ~ "vs Green Energy Adoption"),
                        x = expression("Total CO"[2] ~ "Emissions (Million Tonnes)"),
                        y = expression("Renewable Energy (10"^15 ~ " BTU's)"),
                        fill = "BTU's"
                    ) +
                    theme(
                         panel.grid.major.x = element_line(colour = "grey80"), # Coloured major grid lines
                         panel.grid.major.y = element_line(colour = "grey90"),
                         panel.background = element_rect(fill = "white"),  # White background
                         plot.title = element_text(size = 15, face = "bold")  # Title customization
                    ) 
```


```{r Arrange plots temp & green plots}
# Arrange the two plots vertically
grid.arrange(
            temp_emission_plot, 
            green_energy_plot, 
            nrow = 2  # Arrange the plots in 2 rows
            )
```


Objective: The aim was to explore global CO2 emissions trends and their connections to temperature change and renewable energy adoption.

World Maps:

Purpose: To show total CO2 emissions per country in 2021 and highlight which countries contribute most to emissions.
Reason: To identify major emitters and understand geographical distribution.
Bar Charts:

Purpose: To display the highest and lowest CO2 emitters by year.
Reason: To track which countries have consistently high or low emissions over time, emphasizing trends and changes in global emission patterns.
Line Graphs:

Purpose: To illustrate CO2 emissions trends per industry over the years.
Reason: To understand which industries contribute most to emissions and observe shifts in their impact over time.
Scatter Plots:

Purpose: To analyze the relationships between CO2 emissions and temperature change as well as CO2 emissions and renewable energy use.
Reason: To explore whether countries with higher emissions are experiencing more significant temperature changes and how renewable energy adoption correlates with emissions.



